Does stomatal patterning in amphistomatous leaves minimize the CO$_2$ diffusion path length with leaves of *Arabidopsis thaliana*?

### 2023-11-06 ms/ms.Rmd
Finished paper draft. Added new .bib file to reflect new citations

---
title: Does stomatal patterning in amphistomatous leaves minimize the CO$_2$ diffusion path length with leaves of *Arabidopsis thaliana*?
date: "`r Sys.Date()`"
author:
  - name: Jacob L. Watts
    email: Jacob.Watts-1@colorado.edu
    corresponding_author: yes # Footnote using corresponding_author
    affiliation: 
      - UHM
      - CU
  - name: Graham J. Dow
    email: graham.dow@niab.com
    affiliation: NIAB
  - name: Thomas N. Buckley
    email: tnbuckley@ucdavis.edu
    affiliation: UCD
  - name: Christopher D. Muir
    email: cdmuir@wisc.edu
    affiliation: 
      - UHM
      - UWM
    # footnote: 1 # Footnote using reference
address:
  - code: UHM
    address: School of Life Sciences, University of Hawaiʻi at Mānoa, Honolulu, HI 96822
  - code: CU
    address: Ecology and Evolutionary Biology, Universtity of Colorado, Boulder, CO 80309
  - code: NIAB
    address: Department of Crop Science and Production Systems, NIAB, Cambridge, CB3 0LE, UK
  - code: UCD
    address: Department of Plant Sciences, University of California, Davis, CA 95616
  - code: UWM
    address: Department of Botany, University of Wisconsin, Madison, WI 53706
# footnote:
#   - code: 1
#     text: "Current email address: \\href{mailto:cdmuir@wisc.edu}{cdmuir@wisc.edu}"
abstract:
  - "We will write abstract after Discussion and Conclusions are complete"
keywords:
  - amphistomy
  - Arabidopsis thaliana
  - CO$_2$ diffusion
  - finite element method
  - optimality
  - photosynthesis
  - stomata
fontsize: 12pt
competing_interests: |
  The authors declare no competing interests.
author_contributions: |
  JLW and CDM conceived of the project, analyzed data, and wrote the manuscript. GW provided data. TNB contributed to model development and helped edit the manuscript.
acknowledgements: |
  This work was supported by NSF Award 1929167 to CDM and NSF Award 2307341 to TNB.
bibliography: stomata-spacing.bib
link-citations: yes
output: 
  rticles::oup_article:
    latex_engine: xelatex
    oup_version: 1 # 1 = 2020 CTAN OUP CLS package; 0 = 2009 OUP CLS package
    # journal: "AoB PLANTS"
    document_style: "modern" # Can be contemporary, modern, traditional
    papersize: "large" # Can be large, medium, small
    #citation_package: "default" # Uncomment when using a CSL; default "natbib"
    namedate: TRUE # Set FALSE to use numeric refs; Default FALSE
    number_sections: FALSE # Uncomment to not number sections; default TRUE
    #number_lines: TRUE # Use `lineno` package - Default FALSE
    #number_lines_options: ["mathlines","switch"]  # Options for latex lineno package.
    onecolumn: TRUE # Uncomment for one column format; default FALSE
    extra_dependencies:
      - booktabs # to use with knitr::kable() example below
header-includes:
 - \usepackage[nomarkers,tablesfirst]{endfloat} # For figures and tables at end
 - \usepackage{lineno} # For line numbering
 - \linenumbers # For line numbering
 - \usepackage{hyperref} # For autoref
 - \renewcommand{\figureautorefname}{Fig.}
 - \usepackage[detect-none]{siunitx} # for \numrange{}{}
 - \sisetup{range-phrase = \text{--}}
 - \usepackage{caption} # for \captionsetup{}
 - \usepackage{newunicodechar,graphicx}
 - \DeclareRobustCommand{\okina}{\raisebox{\dimexpr\fontcharht\font`A-\height}{\scalebox{0.8}{`}}}
 - \newunicodechar{ʻ}{\okina}

@@

Stomatal anatomy (e.g. size, density, distribution, and patterning) and movement regulate gas exchange during photosynthesis, namely CO$_2$ assimilation and water loss through transpiration. Since waxy cuticles are mostly impermeable to CO$_2$ and H$_2$O, stomata are the primary entry points through which gas exchange occurs despite making up a small percentage of the leaf area [@lange_responses_1971]. Stomata consist of two guard cells which open and close upon changes in turgor pressure or hormonal cues [@mcadam_linking_2016]. The stomatal pore leads to an internal space known as the substomatal cavity where gases contact the mesophyll. Once in the mesophyll, CO$_2$ diffuses throughout a network of intercellular air space (IAS) and into mesophyll cells where CO$_2$ assimilation ($A$) occurs within the chloroplasts [@lee_diffusion_1964]. Stomatal conductance and transpiration are determined by numerous environmental and anatomical parameters such as vapor pressure deficit (VPD), irradiance, temperature, wind speed, leaf water potential, IAS geometry, mesophyll cell anatomy, and stomatal anatomy.

Many successful predictions about stomata and other leaf traits can be made by hypothesizing that natural selection should optimizes CO$_2$ gain per unit of water loss [@cowan_stomatal_1977; @buckley_optimal_2017; @sperry_predicting_2017]. However, stomatal anatomy may be partially constrained by physical and developments limits on phenotypic expression [@croxdale_stomatal_2000; @harrison_influence_2020; @muir_how_2023]. Sometimes optimization leads to similar phenotypes across many disparate species. For example, almost all stomata follow the one cell spacing rule to maintain proper stomatal functioning [@geisler_oriented_2000; @dow_physiological_2014]; however some species (notably in *Begonia*) appear to benefit from overlapping vapor shells caused by stomatal clustering [@yi_gan_stomatal_2010; @lehmann_effects_2015; @papanatsiou_stomatal_2017]. Stomatal traits also vary adaptively in different environments. Stomatal density positively co-varies with irradiance during leaf development and negatively co-varies with CO$_2$ concentration [@gay_influence_1975; @schoch_dependence_1980; @woodward_stomatal_1987; @royer_stomatal_2001], consistent with optimality predictions. Stomatal size is jointly controlled by genome size, light, and stomatal density [@jordan_environmental_2015]. Size positively co-varies with genome size [@roddy_scaling_2020] and negatively co-varies with stomatal density [@camargo_density_2011]. Total stomatal area (size $\times$ density) is optimized for operational conductance ($g_\text{s,op}$) rather than maximum conductance ($g_\text{s,max}$) such that stomatal apertures are most responsive to changes in the environment at their operational aperture [@franks_physiological_2012; @liu_scaling_2021]. Stomatal aperture can compensate for maladaptive stomatal densities to an extent [@bussis_stomatal_2006], but stomatal density and size ultimately determine a leaf's theoretical $g_\text{s,max}$ [@sack_developmental_2016], which is proportional to $g_\text{s,op}$ [@murray_consistent_2020]. Additionally, low stomatal densities lead to irregular and insufficient CO$_2$ supply and reduced photosynthetic efficiency in areas far from stomata [@pieruschka_lateral_2006; @morison_lateral_2005], while high stomatal densities can reduce water use efficiency (WUE) [@bussis_stomatal_2006] and incur excessive metabolic costs [@deans_optimization_2020]. In most species, stomata occur on the abaxial (usually lower) leaf surface; but amphistomy, the occurrence of stomata on both abaxial and adaxial leaf surfaces, is also prevalent in high light environments with constant or intermittent access to sufficient water [@mott_adaptive_1982; @jordan_using_2014; @muir_light_2018; @drake_two_2019; @muir_is_2019]. Amphistomy effectively halves the CO$_2$ diffusion path length and boundary layer resistance by doubling boundary layer conductance [@parkhurst_adaptive_1978; @harrison_influence_2020; @mott_amphistomy_1991]. Historically, stomatal patterning in dicot angiosperms was thought to be random with an exclusionary distance surrounding each stomate [@sachs_developmental_1974]; however, the developmental controls of stomatal patterning are poorly understood and likely more complex than random development along the leaf surface. @croxdale_stomatal_2000] reviews three developmental theories which attempt to explain stomatal patterning in angiosperms: inhibition, cell lineage, and cell cycle, ultimately arguing for a cell cycle based control of stomatal patterning.

The patterning and spacing of stomata on the leaf affects photosynthesis in $C_3$ leaves by altering the CO$_2$ diffusion path length from stomata to sites of carboxylation in the mesophyll. Maximum photosynthetic rate ($A_\text{max}$) in $C_3$ plants is generally co-limited by biochemistry and diffusion, but modulated by light availability [@parkhurst_intercellular_1990; @manter_ci_2004; @carriqui_diffusional_2015]. Low light decreasing CO$_2$ demand by limiting electron transport rate, leading to relatively high internal CO$_2$ concentration ($C_\text{i}$) and low $A_\text{max}$ [@kaiser_metabolic_2016]. In contrast, well hydrated leaves with open stomata in high light, photosynthesis is often limited by CO$_2$ supply as resistances from the boundary layer, stomatal pore, and mesophyll can result in insufficient CO$C_2$ supply at the chloroplast to maxmimize photosynthesis [@farquhar_biochemical_1980; @lehmeier_cell_2017]. In this study, we focus primarily on how stomatal patterning affects diffusion, ignoring boundary layer and mesophyll resistances.

@@

## Stomatal distribution is nonrandom, but far from ideal

@@

Stomatal spacing on *A. thaliana* leaves partially supports our overall hypothesis that natural selection minimizes the average distance between stomata and chloroplasts, for a given overall stomatal density. There are three nonmutually exclusive hypotheses for why several of our predictions were wrong. First, our predictions are wrong becuase they are based on overly simplistic assumptions about epidermal and mesophyll anatomy. Second, natural selection may be constrained by developmental processes that prevent phenotypes from reaching their adaptive optima. Third, the benefit of some traits may be of too little consequence to result in fitness differences large enough to be respond to selection. We consider the plausibility of these alternative hypotheses below and present ideas for future work to test them.

[this paragraph should discuss why our assumptions may be too simplistic]
We assume an idealized leaf epidermal and mesophyll structure that is homogenous and unconstrained by other tradeoffs. Real leaves not only provide pathways for CO$_2$ diffusion, but must supply water, intercept light, and deter herbivores and pathogens. These competing interests results in nonuniform epidermal and mesophyll structure that could alter our predictions about optimal stomatal spacing. [explain some examples]

[this paragraph should discuss developmental constraint] No developmental pathway exists to ensure the ideal placement of stomata on the leaf. [add more explanation]

[this paragraph should discuss limits on natural selection] The gas exchange model demonstrate little photosynthetic gain from abaxial-adaxial stomatal coordination (\autoref{fig:model_summary}). Even though lateral diffusion may limit photosynthesis [@morison_lateral_2005], the marginal gain from optimally offsetting stomata is not sufficient to generate fitness differences relative to the strength of genetic drift (i.e. the drift-barrier). We can similarly extrapolate that an ideal, equilateral triangular stomatal spacing is only slightly better than a suboptimal overdispered pattern. Explaining these observations as the result of weak selection is in tension with the finding that stomatal size and zone positively covary, which would suggest that small changes in lateral diffusion distance are significant. As described above, the positive correlation between stomatal size and zone may be explained by common developmental processes rather than as an adaptation to maximize CO$_2$ diffusion.

[discuss significance of our study and future directions] Our study corroborates previous studies which demonstrate that stomata are non-randomly distributed along the leaf surface as a result of developmental mechanisms such as spatially biased arrest of stomatal initials [@boetsch_arrest_1995], oriented asymmetric cell division [@geisler_oriented_2000], and cell cycle controls [@croxdale_stomatal_2000]. We do not investigate the potential developmental pathways that influence stomatal dispersion in this study; however, they are important to consider as these pathways could limit plants from reaching the theoretical peak in the adaptive landscape: uniform stomatal dispersion. Instead, as this study suggests, plants may simply compensate with higher stomatal density and by fitting stomatal size to the area that they supply with CO$_2$. To understand why stomata are not ideally dispersed, more modelling should be done to estimate the fitness gain of stomatal dispersion. Additionally, genetic manipulation studies should attempt to create mutants with clustered and ideally dispersed stomata for a comparison of their photosynthetic traits. This could have extremely important implications for maximum assimilation rates in crops as most crop species are grown in high light where CO$_2$ is often limiting. In drought-prone environments, increased stomatal dispersion may increase water use efficiency by reducing the number of stomata needed to achieve the same internal CO$_2$ concentration, $C_\text{i}$.

[Conclusion] Our results suggest that after optimiziing stomatal density and having developmental rules for spacing stomata relatively evenly, there may be limited gains to further optimization. Therefore, developmental constraints may be necessary to make sense of some features of stomatal spacing and distribution.

>>>>

title: Does stomatal patterning in amphistomatous leaves minimize the CO$_2$ diffusion path length with leaves of *Arabidopsis thaliana*?
date: "`r Sys.Date()`"
author:
  - name: Jacob L. Watts
    email: Jacob.Watts-1@colorado.edu
    corresponding_author: yes # Footnote using corresponding_author
    affiliation: 
      - UHM
      - CU
  - name: Graham J. Dow
    email: graham.dow@niab.com
    affiliation: NIAB
  - name: Thomas N. Buckley
    email: tnbuckley@ucdavis.edu
    affiliation: UCD
  - name: Christopher D. Muir
    email: cdmuir@wisc.edu
    affiliation: 
      - UHM
      - UWM
    # footnote: 1 # Footnote using reference
address:
  - code: UHM
    address: School of Life Sciences, University of Hawaiʻi at Mānoa, Honolulu, HI 96822
  - code: CU
    address: Ecology and Evolutionary Biology, University of Colorado, Boulder, CO 80309
  - code: NIAB
    address: Department of Crop Science and Production Systems, NIAB, Cambridge, CB3 0LE, UK
  - code: UCD
    address: Department of Plant Sciences, University of California, Davis, CA 95616
  - code: UWM
    address: Department of Botany, University of Wisconsin, Madison, WI 53706
# footnote:
#   - code: 1
#     text: "Current email address: \\href{mailto:cdmuir@wisc.edu}{cdmuir@wisc.edu}"
abstract:
  - "Photosynthesis is co-limited by multiple factors depending on the plant and its environment. These include biochemical rate limitations, internal and external water potentials, temperature, irradiance, and carbon dioxide (CO2). Amphistomatous leaves have stomata on both abaxial and adaxial leaf surfaces. This feature is considered an adaptation to alleviate CO2 diffusion limitations in productive environments where other factors are not limiting as the diffusion path length from stomate to chloroplast is effectively halved. Plants can also reduce CO2 limitations through other aspects of optimal stomatal anatomy: stomatal density, distribution and patterning, and size. A number of studies have demonstrated that stomata are overdispersed on a single leaf surface; however, much less is known about stomatal anatomy in amphistomatous plants, especially the coordination between leaf surfaces, despite their prevelance in nature and near ubiquity among crop species. Here we use novel spatial statistics based on simulations and photosynthesis modelling to test hypotheses about how amphistomatous plants may optimize CO2 limitations in the model angiosperm Arabidopsis thaliana grown in different light environments. We find that 1) stomata are overdispersed, but not ideally dispersed, on both leaf surfaces across all light treatments; 2) abaxial and adaxial leaf surfaces are independent; and 3) the theoretical improvements to photosynthesis from abaxial-adaxial stomatal coordination are miniscule (~1%) across the range of feasible traitspace. However, we also find that 4) stomatal size is correlated with the mesophyll volume that it supplies with CO2, suggesting that plants may optimize CO2 diffusion limitations through alternative pathways other than ideal, uniform stomatal spacing. We discuss the myriad developmental, physical, and evolutionary constraits which may prohibit plants from reaching the theoretical adaptive peak of uniform stomatal spacing and inter-surface stomatal coordination. These findings contribute to our understanding of amphistomatous plants' exploration of the leaf anatomy adaptive landscape."
keywords:
  - amphistomy
  - Arabidopsis thaliana
  - CO$_2$ diffusion
  - finite element method
  - optimality
  - photosynthesis
  - stomata
fontsize: 12pt
competing_interests: |
  The authors declare no competing interests.
author_contributions: |
  JLW and CDM conceived of the project, analyzed data, and wrote the manuscript. GW provided data. TNB contributed to model development and helped edit the manuscript.
acknowledgements: |
  This work was supported by NSF Award 1929167 to CDM and NSF Award 2307341 to TNB.
bibliography: stomata-spacing_new.bib
link-citations: yes
output: 
  rticles::oup_article:
    latex_engine: xelatex
    oup_version: 1 # 1 = 2020 CTAN OUP CLS package; 0 = 2009 OUP CLS package
    # journal: "AoB PLANTS"
    document_style: "modern" # Can be contemporary, modern, traditional
    papersize: "large" # Can be large, medium, small
    #citation_package: "default" # Uncomment when using a CSL; default "natbib"
    namedate: TRUE # Set FALSE to use numeric refs; Default FALSE
    number_sections: FALSE # Uncomment to not number sections; default TRUE
    #number_lines: TRUE # Use `lineno` package - Default FALSE
    #number_lines_options: ["mathlines","switch"]  # Options for latex lineno package.
    onecolumn: TRUE # Uncomment for one column format; default FALSE
    extra_dependencies:
      - booktabs # to use with knitr::kable() example below
header-includes:
 - \usepackage[nomarkers,tablesfirst]{endfloat} # For figures and tables at end
 - \usepackage{lineno} # For line numbering
 - \linenumbers # For line numbering
 - \usepackage{hyperref} # For autoref
 - \renewcommand{\figureautorefname}{Fig.}
 - \usepackage[detect-none]{siunitx} # for \numrange{}{}
 - \sisetup{range-phrase = \text{--}}
 - \usepackage{caption} # for \captionsetup{}
 - \usepackage{newunicodechar,graphicx}
 - \DeclareRobustCommand{\okina}{\raisebox{\dimexpr\fontcharht\font`A-\height}{\scalebox{0.8}{`}}}
 - \newunicodechar{ʻ}{\okina}

@@

Stomatal anatomy (e.g. size, density, distribution, and patterning) and movement regulate gas exchange during photosynthesis, namely CO$_2$ assimilation and water loss through transpiration. Since waxy cuticles are mostly impermeable to CO$_2$ and H$_2$O, stomata are the primary entry and exit points through which gas exchange occurs despite making up a small percentage of the leaf area [@lange_responses_1971]. Stomata consist of two guard cells which open and close upon changes in turgor pressure or hormonal cues [@mcadam_linking_2016]. The stomatal pore leads to an internal space known as the substomatal cavity where gases contact the mesophyll. Once in the mesophyll, CO$_2$ diffuses throughout a network of intercellular air space (IAS) and into mesophyll cells where CO$_2$ assimilation ($A$) occurs within the chloroplasts [@lee_diffusion_1964]. Stomatal conductance and transpiration are determined by numerous environmental and anatomical parameters such as vapor pressure deficit (VPD), irradiance, temperature, wind speed, leaf water potential, IAS geometry, mesophyll cell anatomy, and stomatal anatomy.

Many successful predictions about stomata and other C3 leaf traits can be made by hypothesizing that natural selection should optimize CO$_2$ gain per unit of water loss for any given set of environmental parameters, including their variability [@cowan_stomatal_1977; @buckley_optimal_2017; @sperry_predicting_2017]. Total stomatal area (size $\times$ density) is optimized for operational conductance ($g_\text{s,op}$) rather than maximum conductance ($g_\text{s,max}$) such that stomatal apertures are most responsive to changes in the environment at their operational aperture [@franks_physiological_2012; @liu_scaling_2021]. Stomatal aperture can compensate for maladaptive stomatal densities to an extent [@bussis_stomatal_2006], but stomatal density and size ultimately determine a leaf's theoretical $g_\text{s,max}$ [@sack_developmental_2016], which is proportional to $g_\text{s,op}$ [@murray_consistent_2020]. Additionally, low stomatal densities lead to irregular and insufficient CO$_2$ supply and reduced photosynthetic efficiency in leaf areas far from stomata [@pieruschka_lateral_2006; @morison_lateral_2005], while high stomatal densities can reduce water use efficiency (WUE) [@bussis_stomatal_2006] and incur excessive metabolic costs [@deans_optimization_2020]. Stomatal density positively co-varies with irradiance during leaf development and negatively co-varies with CO$_2$ concentration [@gay_influence_1975; @schoch_dependence_1980; @woodward_stomatal_1987; @royer_stomatal_2001], consistent with optimality predictions. In most species, stomata occur only on the abaxial (usually lower) leaf surface; but amphistomy, the occurrence of stomata on both abaxial and adaxial leaf surfaces, is also prevalent in high light environments with constant or intermittent access to sufficient water [@mott_adaptive_1982; @jordan_using_2014; @muir_light_2018; @drake_two_2019; @muir_is_2019]. Amphistomy effectively halves the CO$_2$ diffusion path length and boundary layer resistance by doubling boundary layer conductance [@parkhurst_adaptive_1978; @harrison_influence_2020; @mott_amphistomy_1991]. Despite these optimality predictions, stomatal anatomy may be partially constrained by physical and developmental limits on phenotypic expression [@croxdale_stomatal_2000; @harrison_influence_2020; @muir_how_2023].

A number of physical and developmental processes constrain stomatal anatomy traitspace. For example, almost all stomata follow the one cell spacing rule to maintain proper stomatal functioning as guard cell movement requires the rapid exchange of ions with neighboring epidermal cells (i.e. subsidiary cells) [@geisler_oriented_2000; @dow_physiological_2014]. This would prevent stomata from being strongly clustered; however, some species (notably in *Begonia*) appear to benefit from the overlapping vapor shells caused by stomatal clustering in dry environments [@yi_gan_stomatal_2010; @lehmann_effects_2015; @papanatsiou_stomatal_2017]. Historically, stomatal patterning in dicot angiosperms was thought to be random with an exclusionary distance surrounding each stomate [@sachs_developmental_1974]; however, the developmental controls of stomatal patterning are more complex than random development along the leaf surface. @croxdale_stomatal_2000 reviews three developmental theories which attempt to explain stomatal patterning in angiosperms: inhibition, cell lineage, and cell cycle, ultimately arguing for a cell cycle based control of stomatal patterning. @pillitteri_mechanisms_2012 review the short and long distance signalling pathways associated with stomatal spacing and development, which include cell to cell communication and whole plant integration to ensure the proper spacing of stomata across a single leaf surface depending on environmental ques. Much less is known about the development of stomata on the adaxial leaf surface in amphistomatous plants. Stomatal size is additionally constrained by genome size with larger genomes leading to larger minimum guard cell size [@jordan_environmental_2015]. Despite these limitations, ecophysiological theory still predicts optimal stomatal anatomy, the details of which are discussed below.

The patterning and spacing of stomata on the leaf affects photosynthesis in $C_3$ leaves by altering the CO$_2$ diffusion path length from stomata to sites of carboxylation in the mesophyll. Maximum photosynthetic rate ($A_\text{max}$) in $C_3$ plants is generally co-limited by biochemistry and diffusion, but modulated by light availability [@parkhurst_intercellular_1990; @manter_ci_2004; @carriqui_diffusional_2015]. Low light decreases CO$_2$ demand by limiting electron transport rate, leading to relatively high internal CO$_2$ concentration ($C_\text{i}$) and low $A_\text{max}$ [@kaiser_metabolic_2016]. In contrast, well hydrated leaves with open stomata in high light, photosynthesis is often limited by CO$_2$ supply as resistances from the boundary layer, stomatal pore, and mesophyll can result in insufficient CO$C_2$ supply at the chloroplast to maxmimize photosynthesis [@farquhar_biochemical_1980; @lehmeier_cell_2017]. In this study, we focus primarily on how stomatal patterning affects diffusion, ignoring boundary layer and mesophyll resistances.

@@

## Stomatal distribution is non-random, but far from ideal

@@

Stomatal spacing on *A. thaliana* leaves partially supports our overall hypothesis that natural selection minimizes the average distance between stomata and chloroplasts, for a given overall stomatal density. There are three non-mutually exclusive hypotheses for why several of our predictions were wrong. First, our predictions are wrong becuase they are based on overly simplistic assumptions about epidermal and mesophyll anatomy. Second, natural selection may be constrained by developmental processes that prevent phenotypes from reaching their adaptive optima. Third, the benefit of some traits may be of too little consequence to result in fitness differences large enough to respond to selection. We consider the plausibility of these alternative hypotheses below and present ideas for future work to test them.

We assume an idealized leaf epidermal and mesophyll structure that is homogenous and unconstrained by other tradeoffs. Real leaves not only provide pathways for CO$_2$ diffusion, but must supply water, intercept light, and deter herbivores and pathogens. These competing interests result in nonuniform epidermal and mesophyll structure that could alter our predictions about optimal stomatal spacing. In order to maintain consistent leaf water potential across the lamina, stomatal density must be coordinated with vein density [@fiorin_transport_2016]. Thus, stomatal spacing may be optimized not at the interstomatal level, but at a higher level, coordinating water transport and water loss. We also assume a uniform mesophyll; however, in ab-adaxially differentiated leaves, this assumption is overly simplistic. The pallisade mesophyll is more tightly packed than the spongy mesophyll as an adaptation to intercept light efficiently, so lateral diffusion may be more limiting in the adaxial portion of the leaf. This may explain why adaxial leaf surfaces have consistently higher dispersion indices than abaxial surfaces across all light treatments (\autoref{fig:single-surface}). Future gas exchange models should include multiple parameter sets for different strata of mesophyll tissue.

No developmental pathway exists to ensure an idealized placement of stomata on the leaf surface. Rather, stomatal development is a dynamic process that must be plastic to environmental ques. Leaves develop based on short and long distance signalling pathways which relay information about incoming light, humidity, temperature, and surrounding stomata to developing leaf tissues [@pillitteri_mechanisms_2012]. Our results showing an intermediate level of dispersion in stomatal spacing may be best explained by these developmental pathways which ensure the proper spacing of stomata, with an added random effect brought about by the necessity for plasticity in stomatal development (\autoref{fig:single-surface}). However, any nonideal randomness in stomatal spacing may offset by simultaneous and coordinated development of the IAS [@baillie_developmental_2020]. The fact that stomata which supply a greater mesophyll volume tend to be larger suggests that plants may use coordinated development of multiple leaf tissues to negate nonideal stomatal spacing (\autoref{fig:length-area}).

In amphistomatous leaves, ideal stomatal spacing is complicated by a third dimension. Our gas exchange model demonstrates little photosynthetic gain from abaxial-adaxial stomatal coordination (\autoref{fig:model_summary}). Even though lateral diffusion may limit photosynthesis [@morison_lateral_2005], the marginal gain from optimally offsetting stomata is not sufficient to generate fitness differences relative to the strength of genetic drift (i.e. the drift-barrier). We can similarly extrapolate that an ideal, equilateral triangular stomatal spacing is only slightly better than a suboptimal pattern. Any benefit garnered by ideal stomatal spacing may be additionally offset by a cost to developmental flexibility in variable environments [@pillitteri_mechanisms_2012; @baillie_developmental_2020]. Explaining these observations as the result of weak selection is in tension with the finding that stomatal size and zone positively covary, which would suggest that small changes in lateral diffusion distance are significant. As described above, the positive correlation between stomatal size and zone may be explained by common developmental processes rather than as an adaptation to maximize CO$_2$ diffusion. Such flexibility in mature stomate size is an alternate peak in the adaptive landscape which cannot be assessed with our simplistic model assumptions. In any case, there is no evidence for coordinated development of both leaf surfaces, and very little theoretical benefit to photosynthesis, except in marginal circumstances which are exceptionally rare in nature.

Our study corroborates previous studies which demonstrate that stomata are non-randomly distributed along the leaf surface as a result of developmental mechanisms such as spatially biased arrest of stomatal initials [@boetsch_arrest_1995], oriented asymmetric cell division [@geisler_oriented_2000], and cell cycle controls [@croxdale_stomatal_2000]. We do not investigate the potential developmental pathways that influence stomatal dispersion in this study; however, they are important to consider as these pathways could limit plants from reaching a theoretical peak in the adaptive landscape: uniform stomatal dispersion. Instead, as this study suggests, plants may simply compensate with higher stomatal density and by fitting stomatal size to the area that they supply with CO$_2$. To understand why stomata are not ideally dispersed, more modelling (with a more complex set of assumptions including vein density and IAS structure) should be done to estimate the fitness gain of stomatal dispersion. Additionally, genetic manipulation studies should attempt to create mutants with clustered and ideally dispersed stomata for a comparison of their photosynthetic traits. This could have extremely important implications for maximum assimilation rates in crops as most crop species are grown in high light where CO$_2$ is often limiting. In drought-prone environments, increased stomatal dispersion may increase water use efficiency by reducing the number of stomata needed to achieve the same internal CO$_2$ concentration, $C_\text{i}$.

Our results suggest that after optimizing stomatal density and having developmental rules for spacing stomata relatively evenly, there may be limited gains to further optimization. Therefore, developmental constraints may be necessary to make sense of some features of stomatal spacing and distribution. The possibility that ideal stomatal spacing is not the "tallest" fitness peak must also be explored, as stomate size is demonstrated in this study to covary with mesophyll volume supplied with CO$_2$. This may be especially true in highly variable environments or in large tree species with sun and shade leaves where developmental ques may change rapidly. The temporal component, not considered here, could also have significant implications, as CO$_2$ may only be limiting to photosynthesis during short periods when all other conditions are ideal. In these cases, the theoretical benefits of ideal stomatal spacing are further diminished. Future exploration of these competing hypotheses would require more advanced modelling, additional exploration of IAS space development and its effects on gas exchange, both real and modelled, and knowledge about how often the species of interest is CO$_2$ limited across of range of natural settings. Despite these additional considerations, this paper represents an important contribution to understanding the potential drivers of and limitations to stomatal anatomy in amphistomatous plants.
